---
title: "Variant R_t advantage"
author: "Rob Challen"
date: '`r format(Sys.Date(), "%d-%m-%Y")`'
output:
  beamer_presentation:
    slide_level: 2
    theme: "Singapore"
    colortheme: "seagull"
classoption: "aspectratio=169"
fontsize: 10pt
knit: (function(inputFile, encoding,...) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = paste0(here::here("output/",Sys.Date())), output_file=paste0('current-rt-data-',Sys.Date(),'.pdf')) })
# output: 
#   ioslides_presentation:
#     widescreen: true
#     smaller: true
#     css: ~/Git/uk-covid-datatools/vignettes/ioslides.css
#     # self_contained: false
#     # lib_dir: ~/Dropbox/covid19/current-rt/data/libs
# knit: (function(inputFile, encoding,...) {
#   rmarkdown::render(inputFile, encoding = encoding, output_dir = "~/Dropbox/covid19/current-rt/data", output_file=paste0('current-rt-data-',Sys.Date(),'.html')) })
---

## Authors

Robert Challen ^1,2^;  Krasimira Tsaneva-Atanasova ^1,3,5^; Ellen Brooks-Pollock^4^; Leon Danon ^3,5^

1) EPSRC Centre for Predictive Modelling in Healthcare, University of Exeter, Exeter, Devon, UK.
2) Taunton and Somerset NHS Foundation Trust, Taunton, Somerset, UK.
3) The Alan Turing Institute, British Library, 96 Euston Rd, London NW1 2DB, UK.
4) Bristol Medical School, Population Health Sciences, University of Bristol, Bristol, UK
5) Data Science Institute, College of Engineering, Mathematics and Physical Sciences, University of Exeter, Exeter, UK. 

Report: `r Sys.Date()`

## 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = TRUE,
  message = TRUE
)

here::i_am("in-development/rt-advantage.Rmd")
source(here::here("common-setup.R"))
#source(here::here("cron-estimate-rt.R"))## R advantage
```

# mixture distribution combination

```{r}

# TODO make cdfFn a list 
pmixture = function(q, cdfFn, ...) {
  params = rlang::list2(...)
  cdfs = purrr::pmap(params,function(...) return(function(x) cdfFn(x,...)))
  tmp = sapply(cdfs, function(z) z(q))
  if(is.array(tmp)) {
    tmp = rowMeans(tmp)
  } else {
    tmp = mean(tmp)
  }
  return(tmp)
}

# rough test
# pmixture(c(-1), pnorm, mean=c(0,2,4), sd=c(1,2,3))



qmixture = function(p, bounds, cdfFn, ... ) {
  quantGenFn = function(q) return(function(x) pmixture(x, cdfFn, ...)-q)
  estimates = sapply(p, function(x) tryCatch(uniroot(quantGenFn(x), interval=bounds)$root,error = function(e) NaN))
  return(tibble(
    quantile = p,
    estimate = estimates,
    method = "ensemble"
  ))
}

# 
# qmixture(c(0.025,0.5,0.975), c(-1000,1000), pnorm, mean=c(0,2,4), sd=c(1,2,3))

#plot(pmixture(seq(-5,5,length.out = 1001), pnorm, mean=c(0,2,4), sd=c(1,2,3)))

```

# Serial intervals

```{r}

serialInterval = SerialIntervalProvider$generationInterval(dpc)
serialInterval$printSerialInterval()
dfit = serialInterval$dfit
dfitGammas = dfit$discreteProbabilities(q=0:14,summarise = FALSE)
gammaMatrix1 = dfitGammas %>% arrange(bootstrapNumber,value) %>% pull(discreteProbability) %>% matrix(nrow=15,ncol=100)

readr::write_csv(dfit$bootstraps,here::here("output/transmission-advantage/generation-intervals.csv"))

p1 = ggplot(dfitGammas,aes(x=value,y=discreteProbability,group=as.factor(bootstrapNumber)))+geom_line(alpha = 0.2)+ylab("discrete probability")+xlab("day after infection")
p1 %>% standardPrintOutput::saveThirdPageFigure(here::here("output/transmission-advantage/S0-si-paper-gammas"))


```


# Background 

```{r}

infectionEpisodes = dpc$spim$getInfectionEpisodes()
infectionEpisodes = infectionEpisodes %>% rename(specimen_date=earliest_specimen_date)
infectionEpisodes %>% with(table(asymptomatic_indicator,sGene))

dates = infectionEpisodes %>% group_by(sGene) %>% summarise(minDate = min(specimen_date), maxDate = max(specimen_date)) %>% ungroup() %>% summarise(minDate = max(minDate), maxDate = min(maxDate)-4)

symptomaticTimeseries = infectionEpisodes %>% dpc$spim$getLineListIncidence(ageBreaks = NULL,codeTypes = c("CTRY","NHSER"), subgroup = sGene, filterExpr = asymptomatic_indicator != "Y") %>% 
  filter(date <= dates$maxDate & date >= dates$minDate) %>%
  dpc$demog$findDemographics() %>%
  tsp$completeAndRemoveAnomalies()

J = jepidemic::JavaApi$new()

estim = J$CoriEstimator$new(r0Mean = 5,r0SD = 3,maxWindow = 14)
#estim$selectAdaptiveWindow(incidenceSum = 300,minWindow = 7)
estim$selectSpecificWindow(14)
estim$withAdaptivePrior(factor = 1.1)
estim$withInfectivityProfileMatrix(gammaMatrix1)
estim$inMiddleOfTimeseries()
estim$detailedOutput()

symptomaticRtBootstraps = estim$estimateRt(incidence = symptomaticTimeseries %>% ukcovidtools::covidStandardGrouping(), dateColName = "date",incidenceColName = "value.original")
# TODO: investigate factor mapping error fix
# symptomaticRtBootstraps14 = symptomaticRtBootstraps %>% filter(Rt.Window == 14)

estim$collectMixtureQuantiles()
symptomaticRt = estim$estimateRt(incidence = symptomaticTimeseries %>% ukcovidtools::covidStandardGrouping(), dateColName = "date",incidenceColName = "value.original")
symptomaticRt$`Anomaly.R` = symptomaticTimeseries$Anomaly

p1 = tsp$plotRt(symptomaticRt %>% filter(subgroup %in% c("Positive","Negative") & name!="Unknown (England)"), colour = subgroup, dates = "2020-10-01",rtlim = c(0,2.5), ribbons = TRUE)+facet_wrap(vars(name),ncol=4)+scale_color_discrete()+standardPrintOutput::narrowAndTall() + standardPrintOutput::smallLegend()

p1 %>%  standardPrintOutput::saveThirdPageFigure(here::here("output/transmission-advantage/S2-reproduction-number"))


p2 = tsp$plotIncidenceQuantiles(symptomaticTimeseries %>% filter(name!="Unknown (England)"), colour = subgroup, dates = "2020-10-01")+facet_wrap(vars(name),ncol=4)+scale_color_discrete()+standardPrintOutput::narrowAndTall() + standardPrintOutput::smallLegend()+scale_y_continuous(trans="log1p", breaks = ukcovidtools::breaks_log1p())

p2  %>%  standardPrintOutput::saveThirdPageFigure(here::here("output/transmission-advantage/S1-incidence"))
```

# Beta prime mixture methods

```{r}
gammaParamters = 
  symptomaticRtBootstraps %>% filter(subgroup == "Positive") %>% select(code,name,date,Rt.ProfileId,posShape = Rt.Shape, posScale = Rt.Scale) %>%
  left_join(
    symptomaticRtBootstraps %>% filter(subgroup == "Negative") %>% select(code,name,date,Rt.ProfileId,negShape = Rt.Shape, negScale = Rt.Scale),
    by=c("code","name","date","Rt.ProfileId")
  )

rtRatio2 = gammaParamters %>% 
  group_by(code,name,date) %>%
  filter(!is.na(posShape) & !is.na(negShape) & !is.na(posScale) & !is.na(negScale) & negScale > 0) %>%
  group_modify(function(d,g,...) {
    return(qmixture(c(0.025,0.5,0.975), c(-1000,1000), extraDistr::pbetapr, shape1=d$posShape, shape2=d$negShape, scale = d$posScale/d$negScale))
  })

rtRatio2b = gammaParamters %>% 
  group_by(code,name,date) %>%
  filter(!is.na(posShape) & !is.na(negShape) & !is.na(posScale) & !is.na(negScale) & posScale > 0) %>%
  # mutate(
  #   ratio.median = extraDistr::qbetapr(p=0.5,posShape,negShape,posScale/negScale),
  #   ratio.q.025 = extraDistr::qbetapr(p=0.025,posShape,negShape,posScale/negScale),
  #   ratio.q.975 = extraDistr::qbetapr(p=0.975,posShape,negShape,posScale/negScale),
  #   ratio.mean = posMean/negMean
  # )
  group_modify(function(d,g,...) {
    return(qmixture(c(0.025,0.5,0.975), c(-1000,1000), extraDistr::pbetapr, shape1=d$negShape, shape2=d$posShape, scale = d$negScale/d$posScale))
  })


rtRatio3 = bind_rows(
  rtRatio2 %>% pivot_wider(names_from = quantile, values_from = estimate, names_prefix="q.") %>% mutate(advantage="S+ over S-"),
  rtRatio2b %>% pivot_wider(names_from = quantile, values_from = estimate, names_prefix="q.") %>% mutate(advantage="S- over S+")
)

# ggplot(rtRatio3, aes(x=date,y=q.0.5, ymin = q.0.025, ymax = q.0.975)) + geom_ribbon(aes(fill=advantage),alpha=0.4,colour=NA) + geom_line(aes(colour=advantage)) +
#   geom_hline(yintercept=1) +
#   facet_wrap(vars(name))+
#   coord_cartesian(ylim=c(0,2.5))
  
```

# Full bootstrap method

```{r}
## Sensitivity analysis
# TODO: serial interval matrix for gammas with various means with constant shape parameter
# calculate R_t for Spos and Sneg and get detailed output / don't combine bootstraps. 
# 

# psi = ParametricSerialIntervalProvider$new(dpc, "gamma", tibble(param = c("mean","sd"), mean=c(4.2,4.9), sd=c(NA,NA), lower=c(NA,NA), upper=c(NA,NA)))
gammaSamples = 
  symptomaticRtBootstraps %>% filter(subgroup == "Positive") %>% select(code,name,date,Rt.ProfileId,posShape = Rt.Shape, posScale = Rt.Scale) %>%
  left_join(
    symptomaticRtBootstraps %>% filter(subgroup == "Negative") %>% select(code,name,date,Rt.ProfileId,negShape = Rt.Shape, negScale = Rt.Scale),
    by=c("code","name","date","Rt.ProfileId")
  ) %>%
  group_by(code,name,date,Rt.ProfileId) %>%
  filter(!is.na(posShape) & !is.na(negShape) & !is.na(posScale) & !is.na(negScale) & negScale > 0 & posScale > 0) %>%
  group_modify(function(d,g,...) {
    posRt = rgamma(10,shape=d$posShape,scale=d$posScale)
    negRt = rgamma(10,shape=d$negShape,scale=d$negScale)
    ratios = posRt/negRt
    bind_rows(
      tibble(
        ratios = ratios,
        advantage="S+ over S-"
      ),
      tibble(
        ratios = 1/ratios,
        advantage="S- over S+"
      )
    )
  }) 

gammaParamters2 = gammaSamples %>% 
  group_by(code,name,date,advantage) %>%
  summarise(
    q.0.025 = unname(quantile(ratios,0.025)),
    q.0.5 = unname(quantile(ratios,0.5)),
    q.0.975 = unname(quantile(ratios,0.975))
  ) 

# ggplot(gammaParamters2, aes(x=date,y=ratio.median, ymin = ratio.q.025, ymax = ratio.q.975)) + 
#   geom_ribbon(aes(fill=advantage),alpha=0.4,colour=NA) + 
#   geom_line(aes(colour=advantage)) +
#   geom_hline(yintercept=1) +
#   facet_wrap(vars(name))+
#   coord_cartesian(ylim=c(0,2.5))

sensitivityCombination = bind_rows(
  gammaParamters2 %>% mutate(method = "full bootstrap"),
  rtRatio3 %>% mutate(method = "gamma ratio")
)

p1 = ggplot(sensitivityCombination %>% filter(name!="Unknown (England)" & advantage=="S+ over S-" & date > "2020-10-01"), 
       aes(x=date,y=q.0.5, ymin = q.0.025, ymax = q.0.975)) + 
  geom_ribbon(aes(colour=method),fill=NA,linetype="dotted") + geom_line(aes(colour=method)) +
  geom_hline(yintercept=1) +
  facet_wrap(vars(name,method),ncol=4)+
  ylab(latex2exp::TeX("$R_t$ ratio"))+
  coord_cartesian(ylim=c(0,2.5)) +
  standardPrintOutput::narrower() +
  standardPrintOutput::smallLegend()

p1 %>%  standardPrintOutput::saveTwoThirdPageFigure(here::here("output/transmission-advantage/S3-methods-comparison"))

```

# Combination of timeseries and time aggregated R_t advantage


```{r}

b117dates = as.Date(c("2020-12-01","2021-02-01"))
b16172dates = as.Date(c("2021-05-01","2021-06-01"))

rtRatioFromMay = gammaParamters %>% 
  filter(date>b16172dates[1] & date<b16172dates[2]) %>%
  group_by(code,name) %>%
  filter(!is.na(posShape) & !is.na(negShape) & !is.na(posScale) & !is.na(negScale) & posScale > 0) %>%
  # mutate(
  #   ratio.median = extraDistr::qbetapr(p=0.5,posShape,negShape,posScale/negScale),
  #   ratio.q.025 = extraDistr::qbetapr(p=0.025,posShape,negShape,posScale/negScale),
  #   ratio.q.975 = extraDistr::qbetapr(p=0.975,posShape,negShape,posScale/negScale),
  #   ratio.mean = posMean/negMean
  # )
  group_modify(function(d,g,...) {
    return(qmixture(c(0.025,0.5,0.975), c(-1000,1000), extraDistr::pbetapr, shape1=d$posShape, shape2=d$negShape, scale = d$posScale/d$negScale))
  }) %>%
  pivot_wider(names_from = quantile, values_from = estimate, names_prefix="q.") %>% mutate(advantage="B.1.617.2 over B.1.1.7") %>%
  mutate(minDate = b16172dates[1],maxDate = b16172dates[2])

rtRatioJanToMarch = gammaParamters %>% 
  filter(date>b117dates[1] & date<b117dates[2]) %>%
  group_by(code,name) %>%
  filter(!is.na(posShape) & !is.na(negShape) & !is.na(posScale) & !is.na(negScale) & posScale > 0) %>%
  # mutate(
  #   ratio.median = extraDistr::qbetapr(p=0.5,posShape,negShape,posScale/negScale),
  #   ratio.q.025 = extraDistr::qbetapr(p=0.025,posShape,negShape,posScale/negScale),
  #   ratio.q.975 = extraDistr::qbetapr(p=0.975,posShape,negShape,posScale/negScale),
  #   ratio.mean = posMean/negMean
  # )
  group_modify(function(d,g,...) {
    return(qmixture(c(0.025,0.5,0.975), c(-1000,1000), extraDistr::pbetapr, shape1=d$negShape, shape2=d$posShape, scale = d$negScale/d$posScale))
  }) %>%
  pivot_wider(names_from = quantile, values_from = estimate, names_prefix="q.") %>% mutate(advantage="B.1.1.7 over wild-type") %>%
  mutate(minDate = b117dates[1],maxDate = b117dates[2])
```

# Table of estimates

```{r}
rtRatioFromMay %>% bind_rows(rtRatioJanToMarch) %>%
  mutate(value = sprintf("%1.2f (%1.2f \u2014 %1.2f)", q.0.5, q.0.025, q.0.975)) %>%
  filter(name != "Unknown (England)") %>% 
  ungroup() %>%
  select(name,advantage,value) %>%
  pivot_wider(names_from = advantage, values_from=value) %>%
  standardPrintOutput::saveTable(here::here("output/transmission-advantage/T1-advantage-estimate"))
```

# Figure 1 transmission advantage

```{r}
avgAdv = bind_rows(
  rtRatioFromMay, 
  rtRatioJanToMarch
) %>% filter(name != "Unknown (England)")
  
p1 = ggplot(rtRatio3 %>% filter(name != "Unknown (England)"), aes(x=date,y=q.0.5, ymin = q.0.025, ymax = q.0.975)) + 
  geom_ribbon(aes(fill=advantage),alpha=0.4,colour=NA) + 
  geom_line(aes(colour=advantage)) +
  geom_rect(data=avgAdv, mapping=aes(xmin=minDate,xmax=maxDate,ymin = q.0.025, ymax = q.0.975,fill=advantage),inherit.aes = FALSE,alpha=0.3)+
  geom_segment(data=avgAdv, mapping=aes(x=minDate,xend=maxDate,y = q.0.5, yend = q.0.5,colour=advantage), inherit.aes = FALSE)+
  geom_hline(yintercept=1) +
  facet_wrap(vars(name),ncol=4)+
  coord_cartesian(ylim=c(0,2.5),xlim=as.Date(c("2020-10-01",NA)))+
  standardPrintOutput::narrowAndTall()+
  standardPrintOutput::smallLegend()+
  ylab(latex2exp::TeX("$R_t$ ratio"))+
  xlab(NULL)+
  scale_x_date(date_breaks = "2 month",date_labels = "%m-%y")

p1 %>% standardPrintOutput::saveThirdPageFigure(here::here("output/transmission-advantage/Fig1-advantage-timeseries"))

```

# generation time distributions


```{r}

tmp = SerialIntervalProvider$generationInterval(dpc)
tmp$dfit$printDistributionDetail()
shape = tmp$dfit$printDistributionDetail() %>% filter(param=="shape") %>% pull(mean)
mean = tmp$dfit$printDistributionDetail() %>% filter(param=="mean") %>% pull(mean)


gammas = bind_rows(
  lapply(seq(2,7,length.out = 100), function(mean) {
    rate = shape/mean
    cdf = pgamma(0:14, shape, rate)
    dpdf = cdf-lag(cdf,default=0)
    return(tibble(mean=mean,day=0:14,pdf = dpdf))
  })
) %>% arrange(mean,day)

gammaMatrix = gammas %>% arrange(mean,day) %>% pull(pdf) %>% matrix(nrow=15,ncol=100)
profiles = tibble(
  Rt.ProfileId = 0:99,
  gammaMean = gammas$mean %>% unique()
)
  

p1 = ggplot(gammas,aes(x=day,y=pdf,colour=as.factor(mean)))+geom_line()+scale_color_discrete(guide="none")+ylab("discrete probability")+xlab("day after infection")
p1 %>% standardPrintOutput::saveThirdPageFigure(here::here("output/transmission-advantage/S4-explored-gammas"))
```

# Ratio estimates generation time combinations

```{r}
estim2 = J$CoriEstimator$new(r0Mean = 5,r0SD = 5,maxWindow = 14)
estim2$selectSpecificWindow(14)
estim2$withAdaptivePrior(factor = 1.1)
estim2$detailedOutput()
estim2$withInfectivityProfileMatrix(gammaMatrix)

spos = estim2$estimateRt(incidence = symptomaticTimeseries %>% filter(subgroup=="Positive") %>% covidStandardGrouping(), incidenceColName = "value.original", dateColName = "date")
sneg = estim2$estimateRt(incidence = symptomaticTimeseries %>% filter(subgroup=="Negative") %>% covidStandardGrouping(), incidenceColName = "value.original", dateColName = "date")

# filter down to B.1.617.2 emergence dates
spos2 = spos %>% filter(date>b16172dates[1] & date<b16172dates[2]) %>% ungroup() %>% select(code,name,date,Rt.ProfileId,posShape = Rt.Shape, posScale = Rt.Scale) %>% inner_join(profiles, by="Rt.ProfileId") %>% rename(posGammaMean = gammaMean)

sneg2 = sneg %>% filter(date>b16172dates[1] & date<b16172dates[2]) %>% ungroup() %>% select(code,name,date,Rt.ProfileId,negShape = Rt.Shape, negScale = Rt.Scale) %>% inner_join(profiles, by="Rt.ProfileId") %>% rename(negGammaMean = gammaMean)

sposVsneg = spos2 %>% inner_join( sneg2, by=c("code","name","date"), suffix=c(".pos",".neg"))

sposAdvantage = sposVsneg %>% group_by(code,name,Rt.ProfileId.pos,Rt.ProfileId.neg,posGammaMean,negGammaMean) %>%
  group_modify(function(d,g,...) {
    return(qmixture(c(0.025,0.5,0.975), c(-1000,1000), extraDistr::pbetapr, shape1=d$posShape, shape2=d$negShape, scale = d$posScale/d$negScale))
  }) %>%
  pivot_wider(names_from = quantile, values_from = estimate, names_prefix="q.") %>% mutate(advantage="B.1.617.2 over B.1.1.7")

saveRDS(sposAdvantage, here::here("output/transmission-advantage/spos-advantage.RData"))
```

```{r}
p1 = ggplot(sposAdvantage %>% filter(name=="England"), aes(x=posGammaMean, y=negGammaMean))+
  geom_contour(aes(z=q.0.5),binwidth = 0.05)+
  geom_contour(aes(z=q.0.5),breaks = c(1.25,1.5,1.75,2.25,2.5,2.75),colour="magenta")+
  geom_contour(aes(z=q.0.5),breaks = c(1,2,3),colour="red")+
  metR::geom_text_contour(aes(z=q.0.5), stroke = 0.2,binwidth = 0.25)+
  coord_fixed()+
  geom_rect(xmin=4.31, xmax=5.40,ymin=4.31, ymax=5.40,colour="grey50", fill=NA,size=0.5,linetype="dashed")

p2 = ggplot(sposAdvantage %>% filter(name=="England"), aes(x=posGammaMean, y=negGammaMean))+
  geom_contour(aes(z=q.0.025),binwidth = 0.05)+
  geom_contour(aes(z=q.0.025),breaks = c(1.25,1.5,1.75,2.25,2.5,2.75),colour="magenta")+
  geom_contour(aes(z=q.0.025),breaks = c(1,2,3),colour="red")+
  metR::geom_text_contour(aes(z=q.0.025), stroke = 0.2,binwidth = 0.25)+
  coord_fixed()+
  geom_rect(xmin=4.31, xmax=5.40,ymin=4.31, ymax=5.40,colour="grey50",fill=NA,size=0.5,linetype="dashed")

p3 = ggplot(sposAdvantage %>% filter(name=="England"), aes(x=posGammaMean, y=negGammaMean))+
  geom_contour(aes(z=q.0.975),binwidth = 0.05)+
  geom_contour(aes(z=q.0.975),breaks = c(1.25,1.5,1.75,2.25,2.5,2.75),colour="magenta")+
  geom_contour(aes(z=q.0.975),breaks = c(1,2,3),colour="red")+
  metR::geom_text_contour(aes(z=q.0.975), stroke = 0.2,binwidth = 0.25)+
  coord_fixed()+
  geom_rect(xmin=4.31, xmax=5.40,ymin=4.31, ymax=5.40,colour="grey50",fill=NA,size=0.5,linetype="dashed")

p4 = (p1+xlab("S+ mean GT (days)")+ylab("S- mean GT (days)")|((p2+xlab(NULL)+ylab(NULL))/(p3+xlab(NULL)+ylab(NULL))))+patchwork::plot_layout(widths=c(1,0.5))+patchwork::plot_annotation(tag_levels = "A")

p4 %>% standardPrintOutput::saveHalfPageFigure(here::here("output/transmission-advantage/by-generation-time-ENG"))

```

```{r}
p5 = ggplot(sposAdvantage %>% filter(!name %in% c("Unknown (England)","England")), aes(x=posGammaMean, y=negGammaMean))+
  geom_contour(aes(z=q.0.5),binwidth = 0.05)+
  geom_contour(aes(z=q.0.5),breaks = c(1.25,1.5,1.75,2.25,2.5,2.75),colour="magenta")+
  geom_contour(aes(z=q.0.5),breaks = c(1,2,3),colour="red")+
  metR::geom_text_contour(aes(z=q.0.5), stroke = 0.2,binwidth = 0.25)+
  coord_fixed()+facet_wrap(vars(name),ncol=4)+xlab("S+ mean GI (days)")+ylab("S- mean GI (days)")+
  geom_rect(xmin=4.31, xmax=5.40,ymin=4.31, ymax=5.40,colour="grey50",fill=NA,size=0.5,linetype="dashed")

p5 %>% standardPrintOutput::saveHalfPageFigure(here::here("output/transmission-advantage/by-generation-time-NHSER-no-CI"))

```

